Final paper: Towards Proving Termination of \texttt{zelda-mosaic} Algorithms under Conditions on the Input

### 2021-05-03 paper/paper.tex
Update paper.tex

\begin{abstract}
    We report on the development developing matrix algorithms in Coq intended to
    serve proofs about a subset of \matlab\@. Specifically, we are interested in
    a previous set of algorithms that we developed for building mosaics,
    implemented in \matlab\@. We discuss the formalization of the necessary
    matrix algorithms and the relevant \matlab\ structures in Coq~\cite{Coq}, as
    well as what remains to be done to prove the termination of the mosaic
    algorithms under appropriate to-be-determined conditions. We present details
    on the matrix development along with the key theorems for its correctness.
    We describe the challenges we faced in developing the matrix type and its
    operations, providing a glimpse at dependently-typed vectors in Coq. We also
    briefly discuss advanced automation in Coq useful for the matrix
    development.
\end{abstract}

@@

The process of linearizing a matrix is perhaps better explained with a
two-dimensional example. Imagine a general two-dimensional \(3 \times 4\) matrix,
such as the one below, where:
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d \\
        e & f & g & h \\
        i & j & k & l
    \end{bmatrix}
\end{equation*}
To linearize the matrix, we label each value with an increasing index from the
top-left, left-to-right, top-to-bottom. For example:
\begin{equation*}
    \begin{bmatrix}
        a_1 & b_2 & c_3 & d_4 \\
        e_5 & f_6 & g_7 & h_8 \\
        i_9 & j_{10} & k_{11} & l_{12}
    \end{bmatrix}
\end{equation*}
Finally, we create a vector, a one-dimensional matrix, by reading the elements
off in order:
\setcounter{MaxMatrixCols}{20}
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d & e & f & g & h & i & j & k & l
    \end{bmatrix}
\end{equation*}
This process generalizes to higher dimensions. The length of the resulting
vector is always the product of the sizes of the dimensions. To see this, note
that this product counts the number of elements in the matrix and that the
linearized matrix has the same number of elements in a single dimension.

@@

We have informally given pieces of the definition in laying our the relationship
between shape and contents; here, we make the definition more precise. A matrix
is well-formed if, and only if,
\begin{itemize}
    \item the shape is an empty list, and the contents are a single scalar value;
        or
    \item the shape is a non-empty list consisting of a head \(n\) representing
        the length of the first dimension and tail \(t\) representing the shape
        of the subsequent dimensions, and the contents is a \(\func{Matrix}\)
        containing a list with length \(n\), and each element of the list is
        well-formed with respect to \(t\).
\end{itemize}
In Coq, we express this definition via a recursive definition
\(\func{well\_formed'}\) which computes a proposition, or proof-burden, over a
shape and contents. We wrap this up in a definition \(\func{well\_formed}\),
which unpacks a matrix for use with \(\func{well\_formed'}\). With this
definition, we can theorems in terms of matrices that are assumed to be
well-formed or that must be shown to be well-formed.

>>>>

\begin{abstract}
    We report on the development of matrix algorithms in Coq intended to
    serve as proofs about a subset of \matlab\@. Specifically, we are interested in
    a previous set of algorithms that we developed for building mosaics,
    implemented in \matlab\@. We discuss the formalization of the necessary
    matrix algorithms and the relevant \matlab\ structures in Coq~\cite{Coq}, as
    well as what remains to be done to prove the termination of the mosaic
    algorithms under appropriate to-be-determined conditions. We present details
    on the matrix development along with the key theorems for its correctness.
    We describe the challenges we faced in developing the matrix type and its
    operations, providing a glimpse at dependently-typed vectors in Coq. We also
    briefly discuss advanced automation in Coq useful for the matrix
    development.
\end{abstract}

@@

The process of linearizing a matrix is perhaps better explained with a
two-dimensional example. Imagine a general two-dimensional \(3 \times 4\) matrix,
such as the one below:
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d \\
        e & f & g & h \\
        i & j & k & l
    \end{bmatrix}
\end{equation*}
To linearize the matrix, we label each value with an increasing index from the
top-left, left-to-right, top-to-bottom. For example:
\begin{equation*}
    \begin{bmatrix}
        a_1 & b_2 & c_3 & d_4 \\
        e_5 & f_6 & g_7 & h_8 \\
        i_9 & j_{10} & k_{11} & l_{12}
    \end{bmatrix}
\end{equation*}
Finally, we create a vector, a one-dimensional matrix, by reading the elements
off in order:
\setcounter{MaxMatrixCols}{20}
\begin{equation*}
    \begin{bmatrix}
        a & b & c & d & e & f & g & h & i & j & k & l
    \end{bmatrix}
\end{equation*}
This process generalizes to higher dimensions. The length of the resulting
vector is always the product of the sizes of the dimensions. To see this, note
that this product counts the number of elements in the matrix and that the
linearized matrix has the same number of elements in a single dimension.

@@

We have informally given pieces of the definition in laying our the relationship
between shape and contents; here, we make the definition more precise. A matrix
is well-formed if, and only if,
\begin{itemize}
    \item the shape is an empty list, and the contents are a single scalar value;
        or
    \item the shape is a non-empty list consisting of a head \(n\) representing
        the length of the first dimension and tail \(t\) representing the shape
        of the subsequent dimensions, and the contents is a \(\func{Matrix}\)
        containing a list with length \(n\), and each element of the list is
        well-formed with respect to \(t\).
\end{itemize}
In Coq, we express this definition via a recursive definition
\(\func{well\_formed'}\) which computes a proposition, or proof-burden, over a
shape and contents. We wrap this up in a definition \(\func{well\_formed}\),
which unpacks a matrix for use with \(\func{well\_formed'}\). With this
definition, we can write theorems in terms of matrices that are assumed to be
well-formed or that must be shown to be well-formed.
